Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-18 latex/main.tex

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(Figure~\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model \cite{LiStephens2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

Sc2ts uses collection dates to build a time-resolved ARG.
When examining preliminary ARGs,
we found samples that introduced large numbers of mutations (and reversions),
likely due to errors in their collection dates
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1,
which is estimated to have emerged in November, 2021; REF).
To preclude adding such ``time traveling'' samples, we filtered samples by ``HMM cost''.
We define the ``HMM cost'' of a sample to be
the total penalty cost for switches and mismatches (mutations) in its copying path.
We parameterize the LS model such that each mismatch has a penalty cost of one
and a switch has a penalty cost of $k$ (here, set to 4),
which determines how much mutation is favored over recombination
when accounting for a sample (Figure~\ref{fig:methodology}).
If the HMM cost of a sample exceeds a threshold (here, set to 7),
sc2ts does not immediately add the sample to the ARG.

This approach resulted in preliminary ARGs that had noticeably fewer time travelers,
but we found that it also excluded the major ``saltation'' events that characterize many VOCs.
These saltation lineages are characterized by a high number of accumulated mutations,
and are hypothesized to emerge from chronic infections
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}.
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron).
To distinguish time travelers with erroneous dates from true saltation lineages,
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
If on subsequent days these lineages form a sufficiently large group of samples,
sc2ts adds the group back to the ARG.
This ``retrospective sample grouping'' successfully captured the emergence of
many saltation lineages such as Alpha (Figure~\ref{fig:subgraph_alpha}).
The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
in which specific ``seed'' samples were identified for unconditional inclusion
(for details of how seed samples were chosen, see STAR Methods).
First arising in geographic regions with low sampling density,
the initial origins of these VOC are challenging to infer (REF).
With the addition of these manual seeds, we successfully incorporated 
these challenging events (Figures~\ref{fig:subgraph_delta},
\ref{fig:subgraph_ba1_ba2}) and 
all other major lineages (Figure~\ref{fig:samples_per_day}) 
into the ARG.

@@

\subsubsection*{Handling outlier samples}
Although most samples can be explained by a small number of mutations in the ARG,
there are two types of outlier samples which result in unusually high numbers of mutations:
``time travelers'' and ``saltation'' lineages.
Time traveling samples are caused by errors in their collection dates.
For example, if a sample is actually collected 10 months after its reported collection date,
then 20 mutations (instead of, say, 0 or 1) are needed to explain it
using the ARG available on its collection date,
assuming a substitution rate of 2 events per genome per month.
Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
by high numbers of mutations by the time of their detection.
Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
To handle these outlier samples, we introduce three mechanisms:
(1) HMM cost-based rejection,
(2) retrospective sample grouping, and
(3) unconditional seeding.

First, we introduce a rejection mechanism to prevent samples
that are generally poorly explained by the ARG
using the LS model parameterised by a specific value for the recombination penalty ($k$).
For each sample, an HMM cost is calculated based on the number of switches and mismatches
in its most likely copying path under the LS model:
The sum of the penalty cost for the switches ($k$)
and the penalty cost of the mismatches (implicitly, equal to one).
To build the ARG described here, we used $k$ of 4 and a maximum HMM cost of 7.
This combination of parameter values rejected samples which have no switch
but 8 mismatches as well as samples which have one switch and 4 mismatches.
It is important to note that this does not necessarily mean
that samples with more than 8 mismatches
or samples with two or more switches cannot be incorporated into the ARG.
Such samples can be added back using a separate mechanism (see below).
We find that although it effectively removes time travelers with poor sample support (e.g. singletons),
the mechanism can filter samples too aggressively.
For example, in preliminary ARGs, it led to exclusion of major saltation VOCs (e.g. Alpha).

Second, to prevent exclusion of the important saltation VOCs
(and less important but well supported poorly-fit lineages),
we introduce ``retrospective sample grouping''.
The idea is to gather evidence for a lineage within a time window rather than on a single day.
After adding a batch of samples, the samples initially rejected on their collection dates
due to high HMM costs are reconsidered for attachment to the ARG if
(a) they are collected within a retrospective time window (the past 7 days);
(b) they have identical copying paths; and
(c) they have shared immediate reversions
(attaching a new sample to the top or the bottom of a long branch).
Sample groups satisfying these criteria are attached to the ARG if
(a) they have at least 10 samples;
(b) they are collected on at least 3 different dates
(to avoid adding batches of time travelers having the same collection date); and
(c) they have at least two shared ``non-immediate'' reversions
(so as to prevent unrelated time travelers from being grouped together).
Furthermore, phylogenetic tree quality metrics are used to preclude groups of samples
that are unlikely to be closely related.
Specifically, the tree of a sample group was excluded if
(a) it has more than two recurrent mutations or
(b) it has more than 5 mutations per sample on average.

Retrospective sample grouping successfully captured most saltation lineages,
but integrating Delta and Omicron into the ARG required further manual intervention.
To ensure that Delta and Omicron are added, we introduce a third mechanism, ``unconditional seeding''.
The idea is to choose suitable samples belonging to each VOC,
and attach them to the ARG unconditionally (i.e. without filtering using HMM cost).
We sought ``seed'' samples for Delta, Omicron BA.1, and BA.2.
For Delta, we used two sets of seeds:
(1) 8 samples assigned to B.1.617.2, and
(2) 9 additional samples assigned to its sibling lineage B.1.617.1.
Specifically, these samples are the earliest samples present in the Viridian v04 dataset
that were used for the designation of the VOCs.
For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively.
We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
as it could be well explained as as a non-recombinant given the ARG available on its collection date.
For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
which is the earliest BA.2 sample present in the Viridian v04 dataset.
It is dated later than SRR17041376 (the BA.1 seed),
allowing time for the BA.1 lineage to accumulate enough sampled descendants
and to get established in the ARG.
For the full list of seeds, see
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}

In practice, it is difficult to find optimal seeds for saltation lineages,
which ideally would require no reversion to be added to the ARG
and to which sampled descendents would be added with no reversion.
Considering poor early sampling and the presence of time travellers for Delta, Omicron BA.1, and BA.2,
using suboptimal seeds is likely the best that we can do in practice.
The consequence is that there would be some reversions leading to the samples attached to the seed(s).
Nevertheless, it is a better alternative to not using seeds,
as there would be far more reversions (as seen in preliminary ARGs built without seeding).
The caveat of taking this approach is that one must be more careful
when analysing the samples around the seed(s) in the ARG.

>>>>

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(Figure~\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model \cite{LiStephens2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.
[TODO add brief summary of sample filtering and how we handle VOCs, 
refing Methods for details]

@@

\subsubsection*{Inserting saltation lineages}
The HMM cost threshold approach outlined in the previous section is 
effective at removing time-travelling samples, but it also filters out 
important evolutionary events such as 
the major ``saltations'' that characterize many VOCs.
These lineages are characterized by a high number of accumulated mutations,
and are hypothesized to emerge from chronic infections
during which the virus is exposed to individual-specific immune selective
% Other refs here for this Sally?
pressures\cite{Otto2021origins,Markov2023}.
To distinguish time travelers from true saltation lineages,
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
If these samples form a sufficiently plausible group over successive days (under criteria
described in the next paragraph) sc2ts adds the group to the ARG as a 
``retrospective sample group'' using same mechanisms as standard
daily sample clusters.

Distinguishing a truly emerging outbreak from different forms of
correlated error is non-trivial, and sc2ts has 5 parameters to fine
tune the behaviour. Firstly, we require at least 10 samples in the
group to ensure there is sufficient support over a time window of 
7 days. Then, we build a local tree of the samples and consider
some tree metrics to evaluate whether the samples within the 
group are closely related. If we have at least 2 mutations 
shared by all samples, there's at most 2 recurrent mutations
within the group and at most 5 mutations unique to each sample, 
the retrospective group is accepted for inclusion.
Although this process is highly hueristic, it successfully 
captured the majority of saltational lineages
such as Alpha (Figure~\ref{fig:subgraph_alpha}), and the 
vast majority of samples were added following the standard
daily sample grouping mechanisms. 
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_pr_pp_mp_aph_bps_pango_dated_recombinants_report.ipynb
In total, 882,398 daily 
sample groups were added to the ARG, while only 100 were
added via this retrospective grouping mechanism.

The Delta, BA.1, and BA.2 lineages were 
particularly challenging, however, as the initial emergences of 
the saltation lineages were not densely sampled. In this case,
there is little information for the retrospective sampling grouping mechanism
to work with and it is very difficult to distinguish short-term time-travellers
from true outbreak samples. For these cases, we chose specific
high-quality samples with confident dates as ``seeds`` which are 
unconditionally included in the ARG on the specified date.

% We used more seeds, but in practise it doesn't matter and it's just
% complicating the story.
For Delta, we used two seeds: one for B.1.617.1 and 1 samples assigned to
B.1.617.2, [with more details and some references to back up our assertions.
There's some specific info in the Viridian paper on this, I'm sure. 
We should mention long branch splitting here too.]
The resulting subgraph can be see in Figure~\ref{fig:subgraph_delta}.

[TODO revise this paragraph to give some background info, and reference
the Viridian work]
For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively.
We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
as it could be well explained as as a non-recombinant given the ARG available on its collection date.
For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
which is the earliest BA.2 sample present in the Viridian v04 dataset.
It is dated later than SRR17041376 (the BA.1 seed),
allowing time for the BA.1 lineage to accumulate enough sampled descendants
and to get established in the ARG.
See Figure~\ref{fig:subgraph_ba1_ba2} for an overview of the resulting 
subgraph.

With the combination of these mechanisms---gradual accumulation 
of diversity through daily sample groups, retrospective group
inclusion and manual seeding---we successfully incorporated all major
lineages in the dataset into the ARG (Figure~\ref{fig:samples_per_day}).
